Parallel Global Search Algorithm for Optimization of the Kinetic Parameters of Chemical Reactions

### 2021-05-30 2021/RSD_Ufa/RSD2021.tex
RSD2021 final

One of the areas of the Laboratory of Mathematical Chemistry of the Institute of Petrochemistry and Catalysis of the UFIC RAS is the study of chemical reactions by mathematical modeling methods. At the moment, a large number of complex chemical processes have been studied, among  them, the reaction of cycloalumination of alkenes with triethylaluminium into alumacyclopentanes, the reaction of hydroalumination of olefins with organoaluminum compounds, the reaction of alcohols with dimethyl carbonate, etc. The mathematical description of chemical reactions is necessary for the development the mathematical models of chemical kinetics, which are called kinetic models. The kinetic model is the basis for mathematical modeling of chemical reactions and the initial level for mathematical modeling of chemical reactors. When developing a kinetic model of a multi-stage chemical reaction, the inverse problem of chemical kinetics (CK) is solved. The inverse problem of CK is generally called the process of selecting the mechanism of a complex process and determining its quantitative characteristics -- the rate constants of reactions based on experimental data about the dynamics of the process. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem, so the problem of choosing an optimization method is relevant.

@@

The process under study is propane pre-reforming into methane-rich gas over Ni catalyst, which is an industrially important chemical process \cite{USKOV2019126741, Uskov_en}. Pre-reforming of propane was studied over industrial nickel-chromium catalyst at pressure of 1 and 5 bar in the temperature range of 220 -- 380 $^0$C and at flow rates of 4000 and 12000 h$^{-1}$. The experimental data on propane pre-reforming was acquired from our previous work \cite{Uskov_en}.  The reaction scheme consists of two reactions: propane steam conversion and CO$_2$ methanation \cite{Uskov_cat}:
\begin{gather}
C_3H_8 + 6H_2O \rightarrow 10H_2 + 3CO_2 \label{reac_ref}\\
CO_2 + 4H_2 \rightleftharpoons CH_4 + 2H_2O \label{reac_met},
\end{gather}
where the rates of reactions \eqref{reac_ref} -- \eqref{reac_met} are expressed according to the Langmuir-Hinshelwood model \cite{USKOV2019126741}:

@@

\begin{equation}\label{mater_balance}
\begin{cases}
G \dfrac{dy_i}{dl} = \left(\nu_i^{ref} W_{ref} + \nu_i^{met} W_{met} \right) m_i, 
\\
0 \le l \le , i \in \left\lbrace C_3H_8, CH_4, H_2O, H_2, CO_2 \right\rbrace,
\\
l = 0, y_i = y_{i0},
\end{cases}
\end{equation}
where $G$ is a mass flow of the mixture, kg/(m$^2 \cdot$ sec); $y_i$ is a mass fraction of the $i$-th component; $\nu_i$ is a stoichiometric coefficient of the $i$-th component; $m_i$ is a molar mass of the $i$-th component, kg/mol; $l$ is coordinate along the catalytic layer, m; $L$ is a length of the catalytic layer, m. The length of the catalytic layer is 0.008 m. The mathematical model of chemical kinetics problems is a system of differential equations that describes the variations in substance concentrations over time according to the rates of reaction stages. The system of differential equations is a Cauchy problem containing the initial data \cite{ENIKEEV2020123131, Chainikova}. The numerical solving of such a system of equations is a direct problem of chemical kinetics.

>>>>

One of the areas of the Laboratory of Mathematical Chemistry of the Institute of Petrochemistry and Catalysis of the UFIC RAS is the study of chemical reactions by mathematical modeling methods \cite{Zaynullin2020, Zainullin2019, Gubaidullin2014}. At the moment, a large number of complex chemical processes have been studied, among  them, the reaction of cycloalumination of alkenes with triethylaluminium into alumacyclopentanes, the reaction of hydroalumination of olefins with organoaluminum compounds, the reaction of alcohols with dimethyl carbonate, etc. The mathematical description of chemical reactions is necessary for the development the mathematical models of chemical kinetics, which are called kinetic models \cite{Koledina2019}. The kinetic model is the basis for mathematical modeling of chemical reactions and the initial level for mathematical modeling of chemical reactors. When developing a kinetic model of a multi-stage chemical reaction, the inverse problem of chemical kinetics (CK) is solved. The inverse problem of CK is generally called the process of selecting the mechanism of a complex process and determining its quantitative characteristics -- the rate constants of reactions based on experimental data about the dynamics of the process. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem, so the problem of choosing an optimization method is relevant \cite{Enikeeva2021, Enikeeva2020}.

@@

The process under study is propane pre-reforming into methane-rich gas over Ni catalyst, which is an industrially important chemical process \cite{Uskov2019, Uskov2020}. Pre-reforming of propane was studied over industrial nickel-chromium catalyst at pressure of 1 and 5 bar in the temperature range of 220 -- 380 $^0$C and at flow rates of 4000 and 12000 h$^{-1}$. The experimental data on propane pre-reforming was acquired from our previous work \cite{Uskov2020}.  The reaction scheme consists of two reactions: propane steam conversion and CO$_2$ methanation \cite{Uskov2017}:
\begin{gather}
C_3H_8 + 6H_2O \rightarrow 10H_2 + 3CO_2 \label{reac_ref}\\
CO_2 + 4H_2 \rightleftharpoons CH_4 + 2H_2O \label{reac_met},
\end{gather}
where the rates of reactions \eqref{reac_ref} -- \eqref{reac_met} are expressed according to the Langmuir-Hinshelwood model \cite{Uskov2019}:

@@

\begin{equation}\label{mater_balance}
\begin{cases}
G \dfrac{dy_i}{dl} = \left(\nu_i^{ref} W_{ref} + \nu_i^{met} W_{met} \right) m_i, 
\\
0 \le l \le , i \in \left\lbrace C_3H_8, CH_4, H_2O, H_2, CO_2 \right\rbrace,
\\
l = 0, y_i = y_{i0},
\end{cases}
\end{equation}
where $G$ is a mass flow of the mixture, kg/(m$^2 \cdot$ sec); $y_i$ is a mass fraction of the $i$-th component; $\nu_i$ is a stoichiometric coefficient of the $i$-th component; $m_i$ is a molar mass of the $i$-th component, kg/mol; $l$ is coordinate along the catalytic layer, m; $L$ is a length of the catalytic layer, m. The length of the catalytic layer is 0.008 m. The mathematical model of chemical kinetics problems is a system of differential equations that describes the variations in substance concentrations over time according to the rates of reaction stages. The system of differential equations is a Cauchy problem containing the initial data \cite{Enikeev2020, Chainikova2017}. The numerical solving of such a system of equations is a direct problem of chemical kinetics.
